---
title: "A basic glossary for R with LLMs"
author: "NLB"
date: "2025-04-01"
output: html_document
---

Below is a structured glossary of essential concepts and terms needed when teaching/using R as a programming language. These concepts cover foundational elements, syntax, data structures, and advanced functionalities.

# Basic Concepts ---

    Object: Everything in R is treated as an object, such as data sets, variables, functions, or plots. Objects have attributes like class and mode that determine their behavior
    

```{r}
x <- 42  # Creates an object 'x' with the value 42
class(x)  # Output: "numeric"
```


Environment
An environment is a collection of objects (variables, functions) stored in memory. The global environment (.GlobalEnv) is the default workspace in R.



```{r}
ls()  # Lists all objects in the current environment
```


Variable
A variable is a named storage location for values or objects. Variables in R are dynamically typed, meaning their type depends on the assigned value. 
 

```{r}
name <- "John"
age <- 30
print(name)  # Output: "John"
print(age)   # Output: 30
```

Assignment Operator
The <- symbol is the primary operator used to assign values to variables in (e.g., x <- 5)

    
```{r}
    x <- 5  # Assigns value 5 to x
    y = 10  # Alternative assignment syntax
    z <<- 15  # Assigns value globally (across environments)
```


#  Data Structures ------

    Vector
    A vector is the most basic data structure in R, holding elements of the same type (e.g., numeric, character).

    Vector: The most basic data structure in R, holding elements of the same type. Created using c() (e.g., c(1, 2, 3))



Matrix: A two-dimensional array where all elements are of the same type. Created using matrix() or functions like cbind() and rbind()

```{r}
v <- c(1, 2, 3)  # Creates a numeric vector
sum(v)  # Output: 6 (sum of vector elements)
```

Matrix
A matrix is a two-dimensional array where all elements are of the same type.

```{r}
m <- matrix(1:9, nrow = 3, ncol = 3)
print(m)
```

Output:

```{r}
#    [,1] [,2] [,3]
#[1,]    1    4    7
#[2,]    2    5    8
#[3,]    3    6    9
```

Data Frame
A data frame is a tabular data structure where rows represent observations and columns represent variables (which can be of different types). Created using data.frame()


```{r}
df <- data.frame(Name = c("Alice", "Bob"), Age = c(25, 30))
print(df)
```
Output:
    Name Age
  1 Alice 25
  2 Bob    30

List
A list is a flexible data structure that can hold elements of different types (e.g., vectors, data frames) and of different dimensions/size. subsetting with []


```{r}
lst <- list(name = "Alice", age = 25, scores = c(90, 85))
print(lst)
```

Tibble
A tibble is a modern version of a data frame with enhanced usability features from the tidyverse package.

    
```{r}
library(tibble)

tb <- tibble(Name = c("Alice", "Bob"), Age = c(25, 30))
    print(tb)
```

# show its compatibility with tibble only structures

#  Functions and Operators ---

Pipe Operator (%>%): Used to chain functions together for streamlined data manipulation (common in tidyverse)

Logical Operators: Include <, >, ==, !=, &, |, etc., for conditional statements

Arithmetic Operators: Basic mathematical operations like +, -, *, /, etc.



    Function
    A function is a reusable block of code that performs specific tasks. Functions can take arguments and return values.

```{r}
add <- function(a, b) {
    return(a + b)
}

add(3, 5)  # Output: 8
```


# Packages and Libraries ---

    Package: A collection of R functions, data, and documentation bundled together to perform specific tasks (e.g., ggplot2 for plotting)


CRAN (Comprehensive R Archive Network): The primary repository for R packages

    .

Data Manipulation and Visualization

    ggplot2: A popular package for creating complex visualizations based on the grammar of graphics

.

tidyverse: A collection of packages designed for consistent data manipulation and visualization workflows

.

dplyr: Part of tidyverse; provides functions like filter(), select(), and mutate() for efficient data manipulation

    .

# Programming Constructs

    Script: Code written in .R files that can be saved and reused

.

Loop: Constructs like for, while, or repeat used for iterative operations

.

Conditionals: Statements like if and else used for decision-making

    .

Statistical Modeling

    Linear Regression (lm): Used for modeling relationships between variables.

    Logistic Regression (glm): Applied to classification problems where the outcome is binary

.

Machine Learning Models: Algorithms such as decision trees, random forests, and gradient boosting implemented via packages like caret or h2o

    .

# Advanced Concepts ---

    Object-Oriented Programming (OOP):

        S3 System: Informal system supporting single dispatch.

        S4 System: Formal system supporting multiple dispatch and inheritance

    .

Lazy Evaluation: Function arguments are evaluated only when needed during function execution

    .

Reproducible Research

    Emphasizes documenting code and results to enable replication by others. Tools like R Markdown facilitate this process

    Function

```{r}
add <- function(a, b) {
  return(a + b)
}

add(3, 5)  # Output: 8
```

Functions encapsulate reusable code.

Pipe Operator (%>%)

```{r}
library(dplyr)

mtcars %>%
  filter(mpg > 20) %>%
  summarise(mean_mpg = mean(mpg))
```
Pipes streamline data manipulation workflows.

Logical Operators

```{r}
x <- TRUE; y <- FALSE
x & y      # Output: FALSE (AND operator)
x | y      # Output: TRUE (OR operator)
!x         # Output: FALSE (NOT operator)
```
Arithmetic Operators

    
```{r}
    a <- 10; b <- 5
    a + b      # Output: 15 (Addition)
    a - b      # Output: 5 (Subtraction)
    a * b      # Output: 50 (Multiplication)
    a / b      # Output: 2 (Division)
```
Packages and Libraries

    Installing and Using Packages


```{r}
#install.packages("ggplot2")
library(ggplot2)
ggplot(data = mtcars, aes(x = mpg)) + geom_histogram()
```
    Packages extend R’s functionality.

Data Manipulation and Visualization

    Basic Data Manipulation with dplyr

```{r}
library(dplyr)

df <- data.frame(Name = c("Alice", "Bob"), Age = c(25, 30))

df %>%
  filter(Age > 25) %>%
  mutate(NewAge = Age + 5)
```

Visualization with ggplot2

```{r}
    library(ggplot2)

    ggplot(data = mtcars, aes(x = mpg, y = hp)) +
      geom_point() +
      labs(title = "Scatter Plot of MPG vs HP")
```

# Programming Constructs --

    Script Example

```{r}
# Save this code in file script.R and run it using source("script.R")

```{r}
print("Hello World")
```



library(Cairo)

*Loop Example
```{r}
for (i in seq(1,5)) {
  print(i^2) 
}
```


* Conditional Statements
```{r}
    x <- -5
    if (x > 0) {
      print("Positive")
    } else {
      print("Negative")
    }
```


# Statistical Modeling

* Linear Regression Example (lm)
```{r}
model <- lm(mpg ~ hp, data = mtcars)
summary(model)
```

# Advanced Concepts --

*Object-Oriented Programming (OOP) Example (S3)
```{r}
person <- list(name="Alice", age=25)
class(person)
class(person) <- "person"
class(person)
```





